---
title: "Choosing and fitting distributions for a stochastic CRM"
output: 
  bookdown::html_document2:
    base_format: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simple simulation example}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
bibliography: ../inst/REFERENCES.bib # a bibtex style bibliography
biblio-style: apalike
link-citations: true
editor_options: 
  chunk_output_type: console
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE
)
```

```{r setup, results='hide'}
library(collision)
library(Distance) # for distance modelling
library(mc2d)  # for the PERT distribution
library(ggplot2)
library(data.table) # for easy summarising and joins
library(MASS) # for multivariate normal
```

# DRAFT

This is a companion vignette to the simple simulation example. 

The inner functions of the package are designed to take single value inputs. The input functions `define_bird` and `define_turbine` can define single or stochastic inputs. So we need to:

* Define stochastic rules for inputs
  - bird(s): `define_bird()`
  - turbine(s): `define_turbine()`
  - proportion of flights at and below rotor swept height
  - effective detection radius (EDR): obtained from distance modelling [@Buckland2001]
  - encounter rate: `encounter_rate()`
* Sample from the input distributions into the collision calculations to determine the distribution of possible collision outcomes. 

This allows the simulation to more accurately reflect the uncertainty due to both natural variance (e.g. bird wingspan) and measurement uncertainty (e.g. estimated flight heights).

But how do we choose a distribution and fit the parameters for each of these inputs?

Distributions discussed in this vignette:

- Normal distribution (aka Gaussian): typically not appropriate for most values in a CRM because it is not bounded at 0 (i.e. can be negative) which is rarely the case for the measurements used in collision risk modelling.
- Log-normal distribution: a distribution whose logarithm is normally distributed (i.e. $ln(X)$ is normally distributed). It is very useful for modelling positive parameters because it is bounded at 0.
- Beta distribution: useful for modelling proportions because it is bounded between 0 and 1.
- Modified PERT distribution: very useful for modelling from expert elicitation, it takes as parameters a minimum, maximum and mode (or mean). The modified PERT distribution (which is what is used in `mc2d::rpert`) takes a fourth parameter: shape, which determines the peakiness (TODO can't think of the word) of the distribution. This distribution is perfect for use if you're given a range and mean rather than a standard deviation and a mean.

We will also briefly discuss the multi-variate normal and log-normal distributions but for our purposes they're basically the same as their 1D counterparts except that they have covariance as well as variance and will just act as a way to randomly sample a set of correlated values.

There are sort of two kinds of stochastic inputs in this process; there are inputs that should be defined by expert elicitation and those that should be determined from on-site survey results.

## Define inputs

The main inputs that are typically determined by expert elicitation are the bird parameters (from an appropriately qualified ecologist) and the turbine parameters (from the project engineers).

### Turbine model

`collision` requires several turbine model parameters as inputs but majority of these should not vary for a given turbine model (e.g. hub height). However there are a few parameters that can vary for a turbine. 

RPM, tilt and prop operational are the ones that can vary

#### RPM

Pretty sure the project engineers will have a modelled RPM or something similar, so best-case you should use that. If they provide it as a spreadsheet or something and you can't easily fit a pre-defined distribution you can just use the `sample()` function from base R to sample directly from the list of values.

If you really can't get a straight answer or you're in more of the blue-sky phase, perhaps a PERT distribution is the way to go. The min can be the cut-in speed, the max is the maximum operating speed and the mode is the optimal operation speed. The shape parameter should be quite high because the turbines are engineered to spend majority of the time at the optimal speed, so it should be far more common than any other speed.

#### Tilt

Still don't fully understand what tilt is but I assume it has some pretty defined range that is physically valid so probably a PERT would be best?

#### Proportion of year operational

This is typically going to be pretty close to 0.95-1 but there may still be some variation in that range that you would like to capture. As a proportion, this can be easily modelled by a beta distribution.

### Bird species

The exact physical characteristics of the bird species of concern are typically have a lot of variation between individuals, or even for a single individual in the case of speed. 

#### Speed and body length

The bird's physical parameters are the perfect use case for a PERT distribution, things like body length and speed are often given simply as ranges, with little information on the variance or anything like that. 

### Proportions (of day, year and avoidance)

Once again, proportions are often best represented by the beta distribution since it's bounded between 0 and 1. A PERT distribution may also be appropriate if you want to use even tighter bounds, for example, you may want to define that the avoidance rate can never be less than 80% (0.8).

## Survey data

Now we need to determine the input values from the observations. In order to determine the error on the estimate of the parameter we use the bootstrap method to repeatedly re-sample the distribution. This is just an example of how you can obtain the distribution, it is up to the analyst to determine how best to model these parameters.

#### Proportion of flights at rotor swept height

In order to determine the flight flux we need an effective detection height, but unlike with the EDR we can't fit a distance model to the heights because distance modelling relies on the assumption that the birds are uniformly distributed at all distances, which we know is not the case in vertical space (i.e. the density of flights tends to drop off with increasing height). The simplest way to avoid biasing the estimate by artificially inflating or deflating the flux through the turbine is to desktop truncate the observations to the maximum tip height of the turbine and use that as the effective detection height (all the observations can still be used to fit the observer's detection function since it is just used for the horizontal distance correction). Other methods for estimating an effective detection height can be used, but this method is the most straightforward and will work well in almost all cases. Truncating the the maximum turbine height also means that the proportion of flights at rotor swept height (`prop_at_height`) and below rotor swept height (`prop_below_height`) sum to 1. 

The proportion of flights at and below rotor swept height account for the amount of flights at risk of being struck by the blades of the turbine. We only need to calculate a distribution for `prop_below_height` since `prop_at_height = 1 - prop_below_height`. Since the proportion should be bounded at 0 and 1, a beta distribution is a good option.

```{r}
min_rsh1 <- turbine_model1$hh - turbine_model1$rotor_diam * 0.5
max_rsh1 <- turbine_model1$hh + turbine_model1$rotor_diam * 0.5

min_rsh2 <- turbine_model2$hh - turbine_model2$rotor_diam * 0.5
max_rsh2 <- turbine_model2$hh + turbine_model2$rotor_diam * 0.5

## calculate ecdf
# Note - this is a simple example; it's up to the analyst how best to fit the height distribution

prop_below <- function(dt_survey = dt_survey,
                       i = seq_len(nrow(dt_survey)),
                       dt_obs = dt_obs, # filter by max_rsh
                       h = max_rsh){
  
  dt_boot_i <- dt_survey[, .(survey_id = unique(survey_id))][i]
  dt_obs_i <- dt_obs[dt_boot_i, on = .(survey_id), nomatch = 0]
  
  cdf_dat <- ecdf(dt_obs_i$height)
  
  return(cdf_dat(h))
}

# turbine model 1
## bootstrap prop below min RSH
hmin_boot1 <- boot::boot(
  data = dt_survey,
  statistic = prop_below, stype = "i",
  R = 900,
  h = min_rsh1,
  dt_obs = dt_obs[height <= max_rsh1]
)

# some maths to get the beta parameters from the bootstrap
shapes_fun <- function(mean, var){
  a <- mean * (mean * (1 - mean)/var - 1)
  b <- (1 - mean) * (mean * (1 - mean)/var - 1)
  return(c(a, b))
}


h_shapes1 <- shapes_fun(mean = hmin_boot1$t0, var = var(hmin_boot1$t))

prop_below_height1 <- set_random("rbeta", shape1 = h_shapes1[1], shape2 = h_shapes1[2])


# turbine model 2
## bootstrap prop below min RSH
hmin_boot2 <- boot::boot(
  data = dt_survey,
  statistic = prop_below, stype = "i",
  R = 900,
  h = min_rsh2,
  dt_obs = dt_obs[height <= max_rsh2]
)

h_shapes2 <- shapes_fun(mean = hmin_boot2$t0, var = var(hmin_boot2$t))

prop_below_height2 <- set_random("rbeta", shape1 = h_shapes2[1], shape2 = h_shapes2[2])

```

#### Encounter Rate and Effective Detection Radius

This is where we determine the average encounter rate per unit time of survey. That is the average number of individuals observed in each minute of survey (or whatever unit of time you wish to use). This is basically just $\frac{total\ individuals\ observed}{total\ survey\ time}$, however the function also allows for weighting surveys to account for stratification and can apply the Wilson correction [@Wilson1927] if there were no observations. We will also determine the variability in the effective detection radius (EDR) of the observer and how it correlates to the encounter rate (i.e. an observer with a larger EDR should have a higher encounter rate).

For this example, we assume you have done the required distance correction and have arrived at a distance model.

```{r}
summary(ds_raptor)
```

**Notes on Encounter Rate**

As discussed above, we are desktop truncating to the maximum rotor swept height of the turbine, so this encounter rate will be the encounter of birds at at-risk height.

The `encounter_rate()` function takes as input a `df_obs_summary` data.frame. This object has one row per survey and must contain a column named `survey_duration` and a column named `size` which is the **total** number of individuals observed in each **survey** (not observation). Optionally, it can contain a column named `survey_weight` which has the associated relative weights of each survey (to avoid artificially inflating or deflating the encounter rate `sum(survey_weight)` must equal the number of surveys).

First let's make that table (using `data.table`):

```{r}
# turbine model 1
dt_obs_survey1 <- dt_obs[height <= max_rsh1, .("size" = sum(size)), survey_id][dt_survey,
                                                            on = "survey_id"]
# need sum(size) because we want one row per survey (not observation)
setnafill(dt_obs_survey1, cols = c("size"), fill = 0) # set size to 0 for surveys with no observations

# turbine model 2
dt_obs_survey2 <- dt_obs[height <= max_rsh2, .("size" = sum(size)), survey_id][dt_survey,
                                                            on = "survey_id"]
setnafill(dt_obs_survey2, cols = c("size"), fill = 0)
```

Now we can use this table in `encounter_rate()`:

```{r}
obs_fun <- function(dt_survey = dt_survey,
                    i = nrow(dt_survey),
                    dt_obs = dt_obs,
                    ds_model = ds_model,
                    df_obs_summary = dt_obs_survey){
  
  dt_boot_i <- dt_survey[i][order(survey_id)]
  dt_obs_i <- dt_obs[dt_boot_i[, .(survey_id)]
                     , on = .(survey_id)
                     , nomatch = 0]
  df_obs_summary_i <- df_obs_summary[dt_boot_i[, .(survey_id)]
                     , on = .(survey_id)
                     , nomatch = 0]

  ds_boot <- tryCatch(
    ds(data = dt_obs_i[, -c("object")],
              transect = ds_model$call$transect,
              formula = ds_model$call$formula,
              key = ds_model$call$key,
              dht_group = ds_model$call$dht_group,
              order = ds_model$call$order)
    , error = function(e) {
      print(e)
      NULL
    }
  )

  if (is.null(ds_boot)) return(NA)
  edr_boot <- edr_from_distmodel(ds_boot)
  
  er_boot <- encounter_rate(df_obs_summary = df_obs_summary_i,
                            wilson_correction = TRUE # default
                            )
  return(c("EDR" = edr_boot, "encounter_rate" = er_boot))
}

# calling it "flux" because these are the components of the flux calculation
flux_boot1 <- boot::boot(data = dt_survey
                , statistic = obs_fun,
                , sim = "ordinary"
                , stype = "i"
                , R = 200
                , dt_obs = dt_obs
                , df_obs_summary = dt_obs_survey1
                , ds_model = ds_model
  )

flux_boot2 <- boot::boot(data = dt_survey
                , statistic = obs_fun,
                , sim = "ordinary"
                , stype = "i"
                , R = 200
                , dt_obs = dt_obs
                , df_obs_summary = dt_obs_survey2
                , ds_model = ds_model
  )
```

Because the encounter rate and the EDR are correlated, we need to make sure that the randomly sampled values for the simulation match. To do this we use a multivariate distribution. An extra complication is that the encounter rate and the EDR are both positive (i.e. they can't be < 0) so we need a positively bounded distribution. The easiest one to implement here is the log-normal distribution. To use the log-normal distribution without a multi-variate log-normal function, we can input the mean and standard deviation of the log of the variables into `MASS:mvrnorm` and then exponentiate the outputs when we sample from it.

```{r}
# turbine 1
log_mean_EDR1 <- mean(log(flux_boot1$t[,1]))
log_mean_ER1 <- mean(log(flux_boot1$t[,2]))
logcovar1 <- cov(log(flux_boot1$t))
  
obs_flux1 <- set_random("mvrnorm",
                        mu = c("EDR" = log_mean_EDR1, "ER" = log_mean_ER1),
                        Sigma = logcovar1)

## take a look

EDR_hist1 <- sapply(1:1000, FUN = function(i){
  exp(collision::sample_input(obs_flux1)["EDR"])})
hist(EDR_hist1)

encounter_rate_hist1 <- sapply(1:1000, FUN = function(i){
  exp(collision::sample_input(obs_flux1)["ER"])})
hist(encounter_rate_hist1)


# turbine 2
log_mean_EDR2 <- mean(log(flux_boot2$t[,1]))
log_mean_ER2 <- mean(log(flux_boot2$t[,2]))
logcovar2 <- cov(log(flux_boot2$t))
  
obs_flux2 <- set_random("mvrnorm",
                        mu = c("EDR" = log_mean_EDR2, "ER" = log_mean_ER2),
                        Sigma = logcovar2)

## take a look

EDR_hist2 <- sapply(1:1000, FUN = function(i){
  exp(collision::sample_input(obs_flux2)["EDR"])})
hist(EDR_hist2)

encounter_rate_hist2 <- sapply(1:1000, FUN = function(i){
  exp(collision::sample_input(obs_flux2)["ER"])})
hist(encounter_rate_hist2)
```

### Run the simulation

Here we set up a loop to run the simulation. Note that each iteration of the loop is independent to others, meaning that you can wrap the loop inside methods to run the iterations in parallel to save time if you wish.

* Step 0 - Set up simulation parameters
* Step 1 - Sample inputs for bird, turbine, prop at height, prop below height, effective detection radius (EDR) and encounter rate.
* Step 1.5 (optional) - Calculate the `cluster_correction` or any spatial flight corrections. The cluster correction is only needed if the distance between turbines $\leq$ EDR.
* Step 2 - Calculate `turbine_flights_year()` with sampled values - calculate flights through turbine per year
* Step 3 - Run `prob_collision_static()` and `prob_collision_dynamic()` - calculate $P(C|I)$ for one interaction
* Step 4 - Run `n_collisions()` - calculate number of collisions a year

For a breakdown of each step see [the deterministic example](deterministic-example.html).  

```{r}
## inputs needed:
## seed for reproducibility
## iterations - how many runs
## df_turbines  - class `data.frame`
## wte - class `birdInput`
## v90_single - class `turbineInput`
## prop_at_height - class `randomInput`
## prop_below_height - class `randomInput`
## edr - class `randomInput`
## encounter rate - class `randomInput`
## mean height - class `randomInput`


set.seed(1234)
iterations <- 1000

lst_results <- lapply(1:iterations, function(i) {
  df_turbines_results <- df_turbines

  # Step 1 - Sample inputs for bird, turbine, prop at height and prop below height
  bird_i <- sample_input(wte)
  
  
  turbines1_i <- sample_input(turbine_model1)
  prop_below_height1_i <- max( sample_input(prop_below_height1), 0)
  prop_at_height1_i <- 1-prop_below_height1_i
  
  turbines2_i <- sample_input(turbine_model2)
  prop_below_height2_i <- sample_input(prop_below_height2)
  prop_at_height2_i <- 1-prop_below_height2_i
  
  # Step 1.5 calculate the cluster correction
  
  df_turbines_results$cluster_corr <- cluster_correction_a(
    eff_detection_width = 2*edr_i,
    df_turbines = df_turbines_results
  )
  
  # Step 2 calculate flux through turbine per year
  df_turbines_results[df_turbines_results$model == "Turbine 180",
                      "flight_turbine_year"] <- turbine_flights_year(
    survey_type = c("point"),
    encounter_rate = er1_i,
    eff_detection_width = 2*edr_i,
    eff_detection_height = max_rsh1,
    spatial_correction = df_turbines_results[df_turbines_results$model == "Turbine 180",
                      "cluster_corr"],
    rotor_diameter = turbines1_i$rotor_diam,
    hub_height = turbines1_i$hh,
    prop_day = bird_i$prop_day,
    prop_year = bird_i$prop_year
  )
  
  df_turbines_results[df_turbines_results$model == "Turbine 150",
                      "flight_turbine_year"] <- turbine_flights_year(
    survey_type = c("point"),
    encounter_rate = er2_i,
    eff_detection_width = 2*edr_i,
    eff_detection_height = max_rsh2,
    spatial_correction = df_turbines_results[df_turbines_results$model == "Turbine 150",
                      "cluster_corr"],
    rotor_diameter = turbines2_i$rotor_diam,
    hub_height = turbines2_i$hh,
    prop_day = bird_i$prop_day,
    prop_year = bird_i$prop_year
  )

  # Step 3 - Run `p_collisions()` - calculate $P(C|I)$ for one interaction
  df_turbines_results[df_turbines_results$model == "Turbine 180",
                      "p_coll_static"] <- prob_collision_static(
    d_base = turbines1_i$d_base,
    d_rotormin = turbines1_i$d_rotormin,
    d_top = turbines1_i$d_top,
    hh = turbines1_i$hh,
    blade_length = turbines1_i$blade_length,
    max_nac_h = turbines1_i$max_nac_h,
    max_nac_l = turbines1_i$max_nac_l,
    max_width_nacelle = turbines1_i$max_width_nacelle,
    rotor_diam = turbines1_i$rotor_diam,
    tilt_deg = turbines1_i$tilt_deg,
    max_chord = turbines1_i$max_chord,
    min_chord = turbines1_i$min_chord,
    blade_thickness_wide = turbines1_i$blade_thickness_wide,
    blade_thickness_narrow = turbines1_i$blade_thickness_narrow,
    prop_at_height = prop_at_height1_i,
    prop_below_height = prop_below_height1_i
  )
  
  df_turbines_results[df_turbines_results$model == "Turbine 150",
                      "p_coll_static"] <- prob_collision_static(
    d_base = turbines2_i$d_base,
    d_rotormin = turbines2_i$d_rotormin,
    d_top = turbines2_i$d_top,
    hh = turbines2_i$hh,
    blade_length = turbines2_i$blade_length,
    max_nac_h = turbines2_i$max_nac_h,
    max_nac_l = turbines2_i$max_nac_l,
    max_width_nacelle = turbines2_i$max_width_nacelle,
    rotor_diam = turbines2_i$rotor_diam,
    tilt_deg = turbines2_i$tilt_deg,
    max_chord = turbines2_i$max_chord,
    min_chord = turbines2_i$min_chord,
    blade_thickness_wide = turbines2_i$blade_thickness_wide,
    blade_thickness_narrow = turbines2_i$blade_thickness_narrow,
    prop_at_height = prop_at_height2_i,
    prop_below_height = prop_below_height2_i
  )

  df_turbines_results[df_turbines_results$model == "Turbine 180",
                      "p_coll_dyn"] <- prob_collision_dynamic(
    rpm = turbines1_i$rpm, # s_rot,
    blade_length = turbines1_i$blade_length,
    max_width_nacelle = turbines1_i$max_width_nacelle,
    rotor_diam = turbines1_i$rotor_diam,
    blade_thickness_wide = turbines1_i$blade_thickness_wide,
    blade_thickness_narrow = turbines1_i$blade_thickness_narrow,
    hh = turbines1_i$hh,
    bird_length = bird_i$bird_length,
    bird_speed = bird_i$bird_speed,
    prop_at_height = prop_at_height1_i,
    prop_below_height = prop_below_height1_i,
    prop_operational = turbines1_i$prop_operational
  )
  
  df_turbines_results[df_turbines_results$model == "Turbine 150",
                      "p_coll_dyn"] <- prob_collision_dynamic(
    rpm = turbines2_i$rpm, # s_rot,
    blade_length = turbines2_i$blade_length,
    max_width_nacelle = turbines2_i$max_width_nacelle,
    rotor_diam = turbines2_i$rotor_diam,
    blade_thickness_wide = turbines2_i$blade_thickness_wide,
    blade_thickness_narrow = turbines2_i$blade_thickness_narrow,
    hh = turbines2_i$hh,
    bird_length = bird_i$bird_length,
    bird_speed = bird_i$bird_speed,
    prop_at_height = prop_at_height2_i,
    prop_below_height = prop_below_height2_i,
    prop_operational = turbines2_i$prop_operational
  )


  # Step 4 - Run `n_collisions()` - calculate number of collisions a year
  df_turbines_results$n_collision <- n_collision(
    avoidance_rate_static = bird_i$avoidance_static,
    avoidance_rate_dynamic = bird_i$avoidance_dynamic,
    n_flights = df_turbines_results$flight_turbine_year,
    p_coll_static = df_turbines_results$p_coll_static,
    p_coll_dynamic = df_turbines_results$p_coll_dyn
  )

  ## for reporting - optional
  df_turbines_results$iteration <- i

  return(df_turbines_results)
})
```


Now we have the output of each simulation. We can use them to plot out the histogram of collision results. This shows the distribution of the long-term average annual collision rate for the wind farm.

```{r}
lapply(lst_results, function(x) {
  sum(x$n_collision)
}) |>
  unlist() |>
  hist(x = _, xlab = "n_collision", main = "Histogram of collision results")
```

---
## References {-}