---
title: "Choosing distrtibutions for risk modelling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing distrtibutions for risk modelling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(collision)

library(Distance)
library(mc2d)  # for the PERT distribution
```


* Define stochastic rules for inputs
  - bird(s) : `define_bird`
  - turbine(s) : `define_turbine`
  
  
  ### Define turbine model

We set up a turbine - here's an example based basically on a Vestas 90, set up using the `define_turbine` function. These parameters can be defined as probability distributions or as single numbers. 

```{r}
v90_single <- define_turbine(
  model_id = "Vesta V90",
  blade_length = 44,
  blade_thickness_narrow = 0.07,
  blade_thickness_wide = 2.6,
  d_base = 4.15,
  d_rotormin = 3.55,
  d_top = 2.55,
  hh = 65,
  max_chord = 3.51,
  min_chord = 0.39,
  max_nac_h = 4.05,
  max_nac_l = 13.25,
  max_width_nacelle = 3.6,
  rpm = 16.1,
  rotor_diam = NULL,
  tilt_deg = 6,
  prop_operational = 0.98
)
```

### Configure bird inputs

We make a 'bird' object - a list of key parameters. These parameters can be defined as probability distributions or as single numbers. 

```{r}
#TODO issue13 more info on distribuutions

wte <- define_bird(
  species = "Wedge-tailed Eagle",
  bird_length = set_random("rnorm", mean = 0.945, sd = 0.2),
  bird_speed = set_random("rnorm", mean = 17, sd = 3),
  prop_day = set_random("runif", min = 0.48, max = 0.52),
  prop_year = 1,
  avoidance_dynamic = set_random("rpert",
                                 min = 0.88, max = 0.95, 
                                 mode = 0.92, shape =4),
  avoidance_static = 0.9999
)


```


# heights

```{r}
## scratch
## 
## # rotor_diam / 2 is better than using blade length because it
# includes the nacelle size
min_rsh <- v90_single$hh - v90_single$rotor_diam * 0.5
max_rsh <- v90_single$hh + v90_single$rotor_diam * 0.5

## Review the heights

ggplot(data = as.data.frame(df_obs_survey), aes(x = height)) +
  stat_ecdf() +
  geom_vline(xintercept = min_rsh, colour = "red") +
  geom_vline(xintercept = max_rsh, colour = "red")


## calculate ecdf
## Note - this is a simple example; it's up to the analyst how best to
## fit the height distribution

prop_between <- function(dat, i, lwr, upp){

  cdf_dat <- ecdf(dat[i])
  p_low <- cdf_dat(lwr)
  p_high <- cdf_dat(upp)

  return(p_high - p_low)

}

prop_between(df_obs_survey$height, 1:nrow(df_obs_survey),
             lwr = 0, upp = round(min_rsh))


b <- boot::boot(
  data = df_obs_survey$height,
  statistic = prop_between, stype = "i",
  R = 1000,
  lwr = 0, upp = round(min_rsh)
)

boot::boot(
  data = df_obs_survey$height,
  statistic = prop_between, stype = "i",
  R = 1000,
  lwr = round(min_rsh), upp = round(max_rsh)
)

set_random("rnorm", mean = summary(b)$original, sd = summary(b)$bootSE)




prop_below_min <- cdf_height(round(min_rsh))

prop_at_height <- cdf_height(round(max_rsh)) - cdf_height(round(min_rsh))
prop_below_height <- cdf_height(round(min_rsh))


```