---
title: "Deterministic example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Deterministic example}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
template:
  math-rendering: mathml
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(knitr)
library(collision)
library(ggplot2)

# for distance modelling
library(Distance)
```


# Deterministic Example

This is a good example to start with. It models a (fictional) site with 2 proposed turbines, all of the same model, and a (fictional) point count survey of Wedge-tailed Eagle. 
For this example we assume flight density is constant across the site.

We are going to calculate a simple deterministic model output here.  For a stochastic output see the other examples.


## Setup inputs

### Define turbine model

Step one is to set up a turbine - here's an example based basically on a Vestas 90, set up using the `define_turbine` function:

```{r}
v90_single <- define_turbine(
  model_id = "Vesta V90",
  blade_length = 44,
  blade_thickness_narrow = 0.07,
  blade_thickness_wide = 2.6,
  d_base = 4.15,
  d_rotormin = 3.55,
  d_top = 2.55,
  hh = 65,
  max_chord = 3.51,
  min_chord = 0.39,
  max_nac_h = 4.05,
  max_nac_l = 13.25,
  max_width_nacelle = 3.6,
  rpm = 16.1,
  rotor_diam = NULL,
  tilt_deg = 6,
  prop_operational = 0.98
)
```


### Define bird

Now we need a bird (see `define_bird`)

```{r}
wte <- define_bird(
  species = "Wedge-tailed Eagle",
  bird_length = 0.945,
  bird_speed = 17,
  prop_day = 0.5,
  prop_year = 1,
  avoidance_dynamic = 0.90,
  avoidance_static = 0.9999
)
```

### Define a set of turbines

The basic template for the model outputs is a `data.frame` of turbine inputs.  These can be read in from a csv, or you can set them up with a few basic pieces of information and the turbine definition above.  

For each turbine we need a turbine ID, and a location. Location can be NA if you are not including any spatial modelling.


`make_turbine_set` creates the data.frame of turbine inputs, sampled (if needed) and ready for analysis

```{r}
df_turbines <- data.frame(
  turbine_id = c("T01", "T02"),
  lat = c(-32.512, -32.521),
  lon = c(143.441, 143.442)
)
```


### Flight activity rate

Use the provided survey data to fit a (very simple) distance model. **NOTE** Please see [Distance](http://distancesampling.org/R/) and the references therein for a complete primer on doing and interpreting Distance analysis. 

The package dataset:

```{r}
summary(df_obs_survey)
```

is a combined dataset of 100 point count surveys, including the count (`n_individuals`) of raptors (and associated distance at first observation (`distance`)). If none were seen we have an `NA` record for the observation.

It's very important that this data include surveys with no sightings as well as positive detections. 

In addition it must include:

* Distance to the bird at first sighting in metres. If transect sampling, this is usually the perpendicular (right-angle) distance from the transect line.
* Duration of the surveys in minutes
* Count of individuals in each observation (for flock sightings)

We fit a distance model and extract the effective detection radius (EDR)

```{r}
ds_model <- ds(df_obs_survey)

plot(ds_model)

print(edr_from_distmodel(ds_model)) # in metres
print(eda_from_distmodel(ds_model)) # in metres^2
print(w_from_distmodel(ds_model)) # truncation distance / max distance in m
```

From this we can calculate the actitvity rate for the study area. This is in units of 
`flights per sq.km per year`.

Note that I can specify the distance model or the EDA. The results are equivalent

```{r}
activity_rate <- activity_rate_distcorr(
  ds_model = ds_model,
  obs_size = df_obs_survey$n_individuals,
  survey_mins = df_obs_survey$survey_mins,
  wilson_correction = TRUE # default
)
```

In this example, we assume activity rate follows a log normal distribution with mean `activity_rate$mean` and standard deviation `activity_rate$sd`. The units are flights / sqkm / year. It's uncorrected at this stage for daily, monthly or spatial variability.

## Calculate proportion at height

prop_height is either a single value or distribution information which represents the proportion of flights at rotor swept height. An simple, single value example of this is given below

```{r}
turbine_hub_height <- v90_single$hh
turbine_blade_length <- v90_single$blade_length

min_blade_height_m <- turbine_hub_height - turbine_blade_length
max_blade_height_m <- turbine_hub_height + turbine_blade_length

ggplot(df_obs_survey, aes(x = height)) +
  geom_histogram() +
  geom_vline(xintercept = min_blade_height_m, colour = "red") +
  geom_vline(xintercept = max_blade_height_m, colour = "red")

individuals_in_range <- sum(df_obs_survey$n_individuals[df_obs_survey$height > min_blade_height_m &
  df_obs_survey$height < max_blade_height_m])
total_individuals <- sum(df_obs_survey$n_individuals)

# Calculate the proportion
prop_height <- individuals_in_range / total_individuals

print(prop_height)
```

## Modelling

For a deterministic model, we need the following 3 stepss
* Step 1 - Run `n_interaction()` - calculation number of interactions per year
* Step 2 - Run `p_collisions()` - calculate $P(C|I)$ for one interaction
* Step 3 - Run `n_collisions()` - calculate number of collisions a year

### Step 1 - Interactions per turbine per year

This is where we account for

* The size of the turbine
* The proportion of the day active (moderated for the `obs_scale_factor` - see the function help page for more information)
* The proportion of the year active
* The proportion of flights at RSH

```{r}
df_turbines$n_interactions <- n_interaction(
  activity_rate = activity_rate$mean,
  prop_height = 0.75,
  bird_prop_year = wte$prop_year,
  bird_prop_day = wte$prop_day,
  bird_obs_scale_factor = 1.0,
  turb_prop_operational = v90_single$prop_operational,
  turb_blade_length = v90_single$blade_length
)
```

**Note**

* because we are not modelling a spatial activity rate, the turbine interactions are the same for each turbine
* The units for `activity_rate` is flights / sqkm / year and turbine dimensions are in metres
* The units for `n_interactions` is flights / turbine / year. It represents the flights through an area of space equal to the turbine disk at the spatial location of the turbine



### Probability of collision given interaction

Our approach considers both the static ($P(C)_{s}$) and dynamic components ($P(C)_{dyn}$)

Without avoidance, and isolated from the overall collision rate calculation, the total probability of collision, given interaction is the 


$$P(C|I) = P(C|I)_{static} + P(C|I)_{dynamic} $$

because the two options are mutually exclusive.  

Let's calculate the collision probability for a Wedge-tailed Eagle and the V90.


```{r}
df_turbines$p_coll_static <- prob_collision_static(
  d_base = v90_single$d_base,
  d_rotormin = v90_single$d_rotormin,
  d_top = v90_single$d_top,
  hh = v90_single$hh,
  blade_length = v90_single$blade_length,
  max_nac_h = v90_single$max_nac_h,
  max_nac_l = v90_single$max_nac_l,
  max_width_nacelle = v90_single$max_width_nacelle,
  rotor_diam = v90_single$rotor_diam,
  tilt_deg = v90_single$tilt_deg,
  max_chord = v90_single$max_chord,
  min_chord = v90_single$min_chord,
  blade_thickness_wide = v90_single$blade_thickness_wide,
  blade_thickness_narrow = v90_single$blade_thickness_narrow,
  prop_at_height = 0.75,
  prop_below_height = 0.2,
  edr = v90_single$rotor_diam / 2
)

df_turbines$p_coll_dyn <- prob_collision_dynamic(
  rpm = v90_single$rpm, # s_rot,
  blade_length = v90_single$blade_length,
  max_width_nacelle = v90_single$max_width_nacelle,
  rotor_diam = v90_single$rotor_diam,
  blade_thickness_wide = v90_single$blade_thickness_wide,
  blade_thickness_narrow = v90_single$blade_thickness_narrow,
  hh = v90_single$hh,
  bird_length = wte$bird_length,
  bird_speed = wte$bird_speed,
  edr = v90_single$rotor_diam / 2
)

df_turbines$p_coll_dyn + df_turbines$p_coll_static
```

### Step 4 - Total annual collisions (with avoidance)

WE use 99% avoidance rate for static part of turbine and 90% and 95% for dynamic part of turbine.

```{r}
avoidance_rates_static <- 0.99
avoidance_rates_dynamic <- c(0.9, 0.95)

df_avoid <- data.frame(
  species = wte$species,
  avoidance_static = avoidance_rates_static,
  avoidance_dynamic = avoidance_rates_dynamic
)

df_results <- base::merge(
  x = df_turbines,
  y = df_avoid,
  all = TRUE
)


df_results$n_collision <- n_collision(
  avoidance_rate_static = df_results$avoidance_static,
  avoidance_rate_dynamic = df_results$avoidance_dynamic,
  n_interaction_yr = df_results$n_interactions,
  p_coll_static = df_results$p_coll_static,
  p_coll_dynamic = df_results$p_coll_dyn
)

## For a final result, sum all turbines
knitr::kable(
  aggregate(n_collision ~ species + avoidance_dynamic,
    data = df_results, FUN = "sum"
  )
)
```