---
title: "Deterministic example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Deterministic example}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
template:
  math-rendering: mathml
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(knitr)
library(collision)
library(ggplot2)

# for distance modelling
library(Distance)
```


# Deterministic Example

This is a good example to start with. It models a (fictional) site with 2 proposed turbines, all of the same model, and a (fictional) point count survey of Wedge-tailed Eagle. 
For this example we assume flight density is constant across the site.

We are going to calculate a simple deterministic model output here.  For a stochastic output see the other examples.


## Setup inputs

### Define turbine model

Step one is to set up a turbine - here's an example based basically on a Vestas 90, set up using the `define_turbine` function:

```{r}
v90_single <- define_turbine(
  model_id = "Vesta V90",
  blade_length = 44,
  blade_thickness_narrow = 0.07,
  blade_thickness_wide = 2.6,
  d_base = 4.15,
  d_rotormin = 3.55,
  d_top = 2.55,
  hh = 65,
  max_chord = 3.51,
  min_chord = 0.39,
  max_nac_h = 4.05,
  max_nac_l = 13.25,
  max_width_nacelle = 3.6,
  rpm = 16.1,
  rotor_diam = NULL,
  tilt_deg = 6,
  prop_operational = 0.98
)
```


### Define bird

Now we need a bird (see `define_bird`)

```{r}
wte <- define_bird(
  species = "Wedge-tailed Eagle",
  bird_length = 0.945,
  bird_speed = 17,
  prop_day = 0.5,
  prop_year = 1,
  avoidance_dynamic = 0.90,
  avoidance_static = 0.9999
)
```

### Define a set of turbines

The basic template for the model outputs is a `data.frame` of turbine inputs.  These can be read in from a csv, or you can set them up with a few basic pieces of information and the turbine definition above.  

For each turbine we need a turbine ID, and a location. Location can be NA if you are not including any spatial modelling.


`make_turbine_set` creates the data.frame of turbine inputs, sampled (if needed) and ready for analysis

```{r}
df_turbines <- data.frame(
  turbine_id = c("T01", "T02"),
  lat = c(-32.512, -32.521),
  lon = c(143.441, 143.442)
)
```


### Survey, observation and detection radius

Use the provided survey data to fit a (very simple) distance model. **NOTE** Please see [Distance](http://distancesampling.org/R/) and the references therein for a complete primer on doing and interpreting Distance analysis. 

The package dataset:

```{r}
summary(df_obs_survey)
```

is a combined dataset of 100 point count surveys, including the count (`n_individuals`) of raptors (and associated distance at first observation (`distance`)). If none were seen we have an `NA` record for the observation.

It's very important that the field data include surveys with no sightings as well as positive detections. 

In addition it must include:

* Distance to the bird at first sighting in metres. If transect sampling, this is usually the perpendicular (right-angle) distance from the transect line.
* Duration of the surveys in minutes
* Count of individuals in each observation (for flock sightings)

We fit a distance model and extract the effective detection radius (EDR)

```{r}
ds_model <- ds(df_obs_survey)

plot(ds_model)

print(edr_from_distmodel(ds_model)) # in metres
print(w_from_distmodel(ds_model)) # truncation distance / max distance in m
```


## Calculate proportion at height and below height

prop_at_height is either a single value or distribution information which represents the proportion of flights at rotor swept height while prop_below_height is the proportion of flights below rotor swept height.

A simple, single value example of this is given below

```{r}

# rotor_diam / 2 is better than using blade length because it 
# includes the nacelle size 
min_rsh <- v90_single$hh - v90_single$rotor_diam * 0.5
max_rsh <- v90_single$hh + v90_single$rotor_diam * 0.5

# generate a height vector that's one row per observation
height_vec <- rep(df_obs_survey$height, times = df_obs_survey$n_individuals)

## Review the heights

ggplot(data = as.data.frame(height_vec), aes(x = height_vec)) +
  stat_ecdf() +
  geom_vline(xintercept = min_rsh, colour = "red") +
  geom_vline(xintercept = max_rsh, colour = "red")


# calculate ecdf
# Note - this is a simple example; it's up to the analyst how best to 
## fit the height distribution
cdf_height <- ecdf(height_vec)


prop_at_height <- cdf_height(round(max_rsh)) - cdf_height(round(min_rsh))
prop_below_height <- cdf_height(round(min_rsh))


```

* Proportion at height: `r prop_at_height`
* Proportion below height: `r prop_below_height`
* Proportion above height: `r 1.0 - (prop_at_height + prop_below_height)`


## Modelling

For a deterministic model, we need the following 4 steps:

* Step 1 - Run `flight_flux()` and `flux_per_year()` - calculate flux through turbine per year
* Step 2 - Run `prob_interaction_flat()` - calculate $P(I)$
* Step 3 - Run `p_collisions()` - calculate $P(C|I)$ for one interaction
* Step 4 - Run `n_collisions()` - calculate number of collisions a year

### Step 1 - Flux through turbine

This is where we account for

* The size of the turbine
* The proportion of the day active (moderated for the `obs_scale_factor` - see the function help page for more information)
* The proportion of the year active
* The proportion of flights at RSH

We first calculate the flight flux for the turbines. This is in units of 
`flights per minute`. The turbine dimensions are in metres

Note that I can specify the distance model or the edr. The results are equivalent

```{r}
flight_flux_min <- flight_flux(
  ds_model = ds_model,
  obs_size = df_obs_survey$n_individuals,
  survey_mins = df_obs_survey$survey_mins,
  prop_at_height = prop_at_height,
  prop_below_height = prop_below_height,
  rotor_diameter = v90_single$rotor_diam,
  wilson_correction = TRUE # default
)

```

In this example, we assume flight flux follows a log normal distribution with mean `flight_flux_min$mean` and standard deviation `flight_flux_min$sd`. The units are flights / min.

Here we correct it for daily and monthly variability.

```{r}
flight_flux_year <- flux_per_year(
  flux_per_min = flight_flux_min$mean,
  prop_day = wte$prop_day,
  prop_year = wte$prop_year
)
```

* The units for `flight_flux_year` is flights / year

### Probability of interaction

For a flat ultilisation, we can get the probability of interaction using this example

```{r}
prob_interaction <- prob_interaction_flat(num_turbines = nrow(df_turbines))
```


### Step 3 - Probability of collision given interaction

Our approach considers both the static ($P(C)_{s}$) and dynamic components ($P(C)_{dyn}$)

Without avoidance, and isolated from the overall collision rate calculation, the total probability of collision, given interaction is the 


$$P(C|I) = P(C|I)_{static} + P(C|I)_{dynamic} $$

because the two options are mutually exclusive.  

Let's calculate the collision probability for a Wedge-tailed Eagle and the V90 turbine model.


```{r}
df_turbines$p_coll_static <- prob_collision_static(
  d_base = v90_single$d_base,
  d_rotormin = v90_single$d_rotormin,
  d_top = v90_single$d_top,
  hh = v90_single$hh,
  blade_length = v90_single$blade_length,
  max_nac_h = v90_single$max_nac_h,
  max_nac_l = v90_single$max_nac_l,
  max_width_nacelle = v90_single$max_width_nacelle,
  rotor_diam = v90_single$rotor_diam,
  tilt_deg = v90_single$tilt_deg,
  max_chord = v90_single$max_chord,
  min_chord = v90_single$min_chord,
  blade_thickness_wide = v90_single$blade_thickness_wide,
  blade_thickness_narrow = v90_single$blade_thickness_narrow,
  prop_at_height = 0.75,
  prop_below_height = 0.2
)

df_turbines$p_coll_dyn <- prob_collision_dynamic(
  rpm = v90_single$rpm, # s_rot,
  blade_length = v90_single$blade_length,
  max_width_nacelle = v90_single$max_width_nacelle,
  rotor_diam = v90_single$rotor_diam,
  blade_thickness_wide = v90_single$blade_thickness_wide,
  blade_thickness_narrow = v90_single$blade_thickness_narrow,
  hh = v90_single$hh,
  bird_length = wte$bird_length,
  bird_speed = wte$bird_speed,
  prop_at_height = 0.75,
  prop_below_height = 0.2
)


```

### Step 4 - Total annual collisions (with avoidance)


```{r}

df_turbines


df_turbines$n_collision <- n_collision(
  avoidance_rate_static = wte$avoidance_static,
  avoidance_rate_dynamic = wte$avoidance_dynamic,
  flux_yr = flight_flux_year,
  p_interaction = prob_interaction,
  p_coll_static = df_turbines$p_coll_static,
  p_coll_dynamic = df_turbines$p_coll_dyn
)

## For a final result, sum all turbines
sum(df_turbines$n_collision)
```