---
title: "One bird, one turbine example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{One bird, one turbine example}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
template:
  math-rendering: mathml
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(collision)

# for distance modelling
library(Distance)

# for Band Comparison
library(stochLAB)

library(knitr)
```

## The set up

One of the key differences in this package and previous Band-based models is the geometric calculation for the core probability of a collision, given a bird is passing through the turbine zone.

**TODO** smales et all description of the geometry




## A worked example - deterministic

To calculate the probability of collision (of one bird with one turbine) will consider a bird with no avoidance behaviour, and compare the `collision` approach to the basic Band approach,where flights at rotor swept height are assumed to be uniformly dsitributed. 

We've coded an internal function `collision:::p_collision_band` for testing purposes (note the three colons - this function is just for testing) but for the purposes of this vignette we will also use the `stochLAB::get_avg_prob_collision()` package. 


Step one is to set up a turbine - here's an example based basically on a Vestas 90, set up using the `define_turbine` function:

```{r}

str(v90_single) 

```


Now we need a bird (see `define_bird`)

```{r}

str(wte)


```


## One bird one turbine

Our approach considers both the static ($P(C)_{s}$) and dynamic components ($P(C)_{dyn}$)

Without avoidance, and isolated from the overall collision rate calculation, the total probability of collision, given interaction is the 


$$P(C|I) = P(C|I)_{static} + P(C|I)_{dynamic} $$

because the two options are mutually exclusive.  

Let's calculate the collision probability for a Wedge-tailed Eagle and the V90.


```{r}

p_coll_static <- prob_collision_static(
  d_base = v90_single$d_base,
  d_rotormin = v90_single$d_rotormin,
  d_top = v90_single$d_top,
  hh = v90_single$hh,
  blade_length = v90_single$blade_length,
  max_nac_h = v90_single$max_nac_h,
  max_nac_l = v90_single$max_nac_l,
  max_width_nacelle = v90_single$max_width_nacelle,
  rotor_diam = v90_single$rotor_diam,
  tilt_deg = v90_single$tilt_deg,
  max_chord = v90_single$max_chord,
  min_chord = v90_single$min_chord,
  blade_thickness_wide = v90_single$blade_thickness_wide,
  blade_thickness_narrow = v90_single$blade_thickness_narrow,
  prop_at_height = 0.75,
  prop_below_height = 0.2, 
  edr = v90_single$rotor_diam/2
)

p_coll_dyn <- prob_collision_dynamic(
  rpm = v90_single$rpm, #s_rot,
  blade_length = v90_single$blade_length,
  max_width_nacelle = v90_single$max_width_nacelle,
  rotor_diam = v90_single$rotor_diam,
  blade_thickness_wide = v90_single$blade_thickness_wide,
  blade_thickness_narrow = v90_single$blade_thickness_narrow,
  hh = v90_single$hh,
  bird_length = wte$bird_length,
  bird_speed = wte$bird_speed,
  edr = v90_single$rotor_diam/2
)

p_coll_dyn + p_coll_static

```

### Hang on - what's that `edr`?

Note we set our effective detection radius (`edr`) to half the rotor diameter. Normally this parameter is calculated via distance sampling (TODO BUCKLAND REF). As the effective detection increases, this probability of collision given interaction also decreases. 

But don't stress! 

As the effective detection radius increases, this also influences the flight density calculation. We will come back to this in a different vignette. For now, we want to compare with Band (as far as possible) so let's use a default thats the flux window associated with the turbine itself.

### Back to Business - whats' Band say?

```{r}

# stochLAB::get_avg_prob_collision(
#   flight_speed = wte$bird_speed, #m/s
#   body_lt = wte$bird_length,
#   wing_span = 2.12,
#   prop_upwind = 0.5,
#   flap_glide = 1,
#   rotor_speed = v90_single$rpm  ,
#   rotor_radius = v90_single$rotor_diam / 2.0,
#   blade_width = v90_single$blade_thickness_wide,
#   blade_pitch =  30 * pi / 180, #In radian
#   n_blades = 3,
#   chord_prof = chord_prof_5MW
# )


# using the 2007 smaller profile from the spreadsheet
collision:::p_collision_band(
  max_chord = v90_single$blade_thickness_wide,
  pitch_deg = 30, 
  rotor_diam = v90_single$rotor_diam,
  rpm = v90_single$rpm,
  bird_length = wte$bird_length,
  bird_wing_span = 2.12,
  bird_flapping = 0,
  bird_speed = wte$bird_speed
)

```

### Calculating flight activity rate

Use the provided survey data to fit a (very simple) distance model. **NOTE** Please see [Distance](http://distancesampling.org/R/) and the references therein for a complete primer on doing and interpreting Distance analysis. 

The package dataset:

```{r}
summary(df_obs_survey)
```

is a combined dataset of 100 point count surveys, including the count (`n_individuals`) of raptors (and associated distance at first observation (`distance`)). If none were seen we have an `NA` record for the observation.

It's very important that this data include surveys with no sightings as well as positive detections. 

In addition it must include:

* Distance to the bird at first sighting in metres. If transect sampling, this is usually the perpendicular (right-angle) distance from the transect line.
* Duration of the surveys in minutes
* Count of individuals in each observation (for flock sightings)

We fit a distance model and extract the effective detection radius (EDR)

```{r}
ds_model <- ds(df_obs_survey)

plot(ds_model)

print(edr_from_distmodel(ds_model)) # in metres
print(eda_from_distmodel(ds_model)) # in metres^2
print(w_from_distmodel(ds_model)) # truncation distance / max distance in m
```

From this we can calculate the actitvity rate for the study area. This is in units of 
`flights per sq.km per year`.

Note that I can specify the distance model or the EDA. The results are equivalent

```{r}

activity_rate <- activity_rate_distcorr(
  ds_model = ds_model,
  obs_size = df_obs_survey$n_individuals,
  survey_mins = df_obs_survey$survey_mins,
  wilson_correction = TRUE # default
)
```

In this example, we assume activity rate follows a log normal distribution with mean `activity_rate$mean` and standard deviation `activity_rate$sd`. The units are flights / sqkm / year. It's uncorrected at this stage for daily, monthly or spatial variability.


### Iinteractions per turbine per year

This is where we account for

* The size of the turbine
* The proportion of the day active (moderated for the `obs_scale_factor` - see the function help page for more information)
* The proportion of the year active
* The proportion of flights at RSH

```{r}
n_interactions_year <-  n_interaction(
  activity_rate = activity_rate$mean,
  prop_height = 0.75,
  bird_prop_year = wte$prop_year, 
  bird_prop_day = wte$prop_day,
  bird_obs_scale_factor = 1.0, 
  turb_prop_operational = v90_single$prop_operational, 
  turb_blade_length = v90_single$blade_length
)

```

**Note**

* because we are not modelling a spatial activity rate, the turbine interactions are the same for each turbine
* The units for `activity_rate` is flights / sqkm / year and turbine dimensions are in metres
* The units for `n_interactions` is flights / turbine / year. It represents the flights through an area of space equal to the turbine disk at the spatial location of the turbine

### Total annual collisions (with avoidance)

Model 90% avoidance

```{r}
avoidance_rate <- 0.9

n_collision_year <- n_collision(
  avoidance_rate = avoidance_rate,
  n_interaction_yr = n_interactions_year,
  p_collision = p_coll_dyn + p_coll_static
  
)

print(n_collision_year)
```