---
title: "One bird, one turbine example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{One bird, one turbine example}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
template:
  math-rendering: mathml
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(collision)

# for Band Comparison
library(stochLAB)

library(knitr)
```

## The set up

One of the key differences in this package and previous Band-based models is the geometric calculation for the core probability of a collision, given a bird is passing through the turbine zone.

**TODO** smales et all description of the geometry




## A worked example - deterministic

To calculate the probability of collision (of one bird with one turbine) will consider a bird with no avoidance behaviour, and compare the `collision` approach to the basic Band approach,where flights at rotor swept height are assumed to be uniformly dsitributed. 

We've coded an internal function `collision:::p_collision_band` for testing purposes (note the three colons - this function is just for testing) but for the purposes of this vignette we will also use the `stochLAB::get_avg_prob_collision()` package. 


Step one is to set up a turbine - here's an example based basically on a Vestas 90, set up using the `define_turbine` function:

```{r}

str(v90_single) 

```


Now we need a bird (see `define_bird`)

```{r}

str(wte)


```


## One bird one turbine

Our approach considers both the static ($P(C)_{s}$) and dynamic components ($P(C)_{dyn}$)

Without avoidance, and isolated from the overall collision rate calculation, the total probability of collision, given interaction is the 


$$P(C|I) = P(C|I)_{static} + P(C|I)_{dynamic} $$

because the two options are mutually exclusive.  

Let's calculate the collision probability for a Wedge-tailed Eagle and the V90.


```{r}

p_coll_static <- prob_collision_static(
  d_base = v90_single$d_base,
  d_rotormin = v90_single$d_rotormin,
  d_top = v90_single$d_top,
  hh = v90_single$hh,
  blade_length = v90_single$blade_length,
  max_nac_h = v90_single$max_nac_h,
  max_nac_l = v90_single$max_nac_l,
  max_width_nacelle = v90_single$max_width_nacelle,
  rotor_diam = v90_single$rotor_diam,
  tilt_deg = v90_single$tilt_deg,
  max_chord = v90_single$max_chord,
  min_chord = v90_single$min_chord,
  blade_thickness_wide = v90_single$blade_thickness_wide,
  blade_thickness_narrow = v90_single$blade_thickness_narrow,
  prop_at_height = 0.75,
  prop_below_height = 0.2
)

p_coll_dyn <- prob_collision_dynamic(
  rpm = v90_single$rpm, #s_rot,
  blade_length = v90_single$blade_length,
  max_width_nacelle = v90_single$max_width_nacelle,
  rotor_diam = v90_single$rotor_diam,
  blade_thickness_wide = v90_single$blade_thickness_wide,
  blade_thickness_narrow = v90_single$blade_thickness_narrow,
  hh = v90_single$hh,
  bird_length = wte$bird_length,
  bird_speed = wte$bird_speed,
  prop_at_height = 0.75,
  prop_below_height = 0.2
)

p_coll_dyn + p_coll_static

```

### Hang on - what's that `edr`?

Note we set our effective detection radius (`edr`) to half the rotor diameter. Normally this parameter is calculated via distance sampling (TODO BUCKLAND REF). As the effective detection increases, this probability of collision given interaction also decreases. 

But don't stress! 

As the effective detection radius increases, this also influences the flight density calculation. We will come back to this in a different vignette. For now, we want to compare with Band (as far as possible) so let's use a default thats the flux window associated with the turbine itself.

### Back to Business - whats' Band say?

```{r}

stochLAB::get_avg_prob_collision(
  flight_speed = wte$bird_speed, #m/s
  body_lt = wte$bird_length,
  wing_span = 2.12,
  prop_upwind = 0.5,
  flap_glide = 1,
  rotor_speed = v90_single$rpm  ,
  rotor_radius = v90_single$rotor_diam / 2.0,
  blade_width = v90_single$blade_thickness_wide,
  blade_pitch =  30 * pi / 180, #In radian
  n_blades = 3,
  chord_prof = chord_prof_5MW
)


# using the 2007 smaller profile from the spreadsheet
collision:::p_collision_band(
  max_chord = v90_single$blade_thickness_wide,
  pitch_deg = 30, 
  rotor_diam = v90_single$rotor_diam,
  rpm = v90_single$rpm,
  bird_length = wte$bird_length,
  bird_wing_span = 2.12,
  bird_flapping = 0,
  bird_speed = wte$bird_speed
)

```