---
title: "One bird, one turbine example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{One bird, one turbine example}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(collision)

# for Band Comparison
library(stochLAB)

library(knitr)
```

## The set up

One of the key differences in this package and previous Band-based models is the geometric calculation for the core probability of a collision, given a bird is passing through the turbine zone.

**TODO** smales et all description of the geometry

## A worked example - deterministic

Step one is to set up a turbine - here's an example based basically on a Vestas 90, set up using the `define_turbine` function:

```{r}

str(v90_single) 

```


Now we need a bird (see `define_bird`)

```{r}




```









```{r}

# 
# lst_birds <- list(
#   "raptor" = define_bird(
#     species = 'biggus raptorus',
#     bird_length = set_random("rnorm", mean = 90, sd = 5), # (metres)
#     bird_speed = set_random("rnorm", mean = 16.67, sd = 3) # m/s
#   ),
#   "little_peep" = define_bird(
#     species = 'littlus birdii',
#     bird_length = set_random("rnorm", mean = 40, sd = 2), # (metres)
#     bird_speed = set_random("rnorm", mean = 21, sd = 3) # m/s
#   )
# )




```


