---
title: "Simple simulation example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simple simulation example}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE
)
```

```{r setup, results='hide'}
library(collision)
library(Distance)
library(mc2d)  # for the PERT distribution
library(ggplot2)
library(data.table)
```


# Stochastic Example


Flow is: define turbine and bird inputs, get distributions for observations (edr, height and encounter rate), put into the big loop.


## Define inputs

### Turbine model

```{r}
v90_single <- define_turbine(
  model_id = "Vesta V90",
  blade_length = 44,
  blade_thickness_narrow = 0.07,
  blade_thickness_wide = 2.6,
  d_base = 4.15,
  d_rotormin = 3.55,
  d_top = 2.55,
  hh = 65,
  max_chord = 3.51,
  min_chord = 0.39,
  max_nac_h = 4.05,
  max_nac_l = 13.25,
  max_width_nacelle = 3.6,
  rpm = set_random("rnorm", mean = 16.1, sd = 2),
  rotor_diam = NULL,
  tilt_deg = 6,
  prop_operational = 0.98
)
```

### Bird species

```{r}
#TODO issue13 more info on distributions

wte <- define_bird(
  species = "Wedge-tailed Eagle",
  bird_length = set_random("rnorm", mean = 0.945, sd = 0.2) ,
  bird_speed = set_random("rnorm", mean = 17, sd = 3),
  prop_day = set_random("runif", min = 0.48, max = 0.52),
  prop_year = 1,
  avoidance_dynamic = set_random("rpert",
                                 min = 0.88, max = 0.95, 
                                 mode = 0.92, shape = 4),
  avoidance_static = 0.9999
)


```

Note: the Beta PERT distribution (`rpert`) is a good distribution to use if you only have the mode (or mean) and the minimum and maximum parameters for a species (eg. mean body length = 0.95m, range: 0.6-1.2m). See _vignette we haven't made yet_ to see how to fit it.

### Survey data

Now we need to determine the input values from the observations.

#### Effective Detection Radius

```{r}
summary(ds_model)
```


```{r message = FALSE}

edr_fun <- function(d = df_obs$survey_id
                    , i = seq_len(nrow(df_obs))
                    , dat = df_obs){

  df_boot_i <- dat[dat$survey_id %in% d[i],]

  ds_CI <- tryCatch(
    ds(data = df_boot_i,
       transect = "point",
       formula = ~ 1,
       key = "hn",
       dht_group = TRUE,
       nadj = 0)
    , error = function(e) {
      print(e)
      NULL
    }
  )

  if (is.null(ds_CI)) return(NA)
  edr_boot <- edr_from_distmodel(ds_CI)
  return(edr_boot)
}

ds_boot <- boot::boot(data = df_obs_survey$survey_id
                , statistic = edr_fun,
                , sim = "ordinary"
                , stype = "i"
                , R = 150
)


## Define EDR
edr <- set_random("rnorm",
                  mean = ds_boot$t0,
                  sd = sd(ds_boot$t, na.rm = TRUE)
                  )

## take a look
EDR_hist <- sapply(1:1000, FUN = function(i) collision::sample_input(edr) )
hist(EDR_hist)
```

#### Flight heights

Mean flight height

```{r}

sample_mean <- function(x, i) {
  return(mean(x[i]))
}

h_boot <- boot::boot(data = df_obs$height*df_obs$size # height scaled by number of individuals
                , statistic = sample_mean,
                , sim = "ordinary"
                , stype = "i"
                , R = 150
)


## Define mean height
mean_h <- set_random("rnorm",
                  mean = h_boot$t0,
                  sd = sd(h_boot$t, na.rm = TRUE)
                  )

## take a look
mean_height_hist <- sapply(1:1000, FUN = function(i) collision::sample_input(mean_h) )
hist(mean_height_hist)
```

Proportion of flights at and below rotor swept height

```{r}
min_rsh <- v90_single$hh - v90_single$rotor_diam * 0.5
max_rsh <- v90_single$hh + v90_single$rotor_diam * 0.5

## calculate ecdf
## Note - this is a simple example; it's up to the analyst how best to
## fit the height distribution

prop_between <- function(dat, i, lwr, upp){

  cdf_dat <- ecdf(dat[i])
  p_low <- cdf_dat(lwr)
  p_high <- cdf_dat(upp)

  return(p_high - p_low)

}

## bootstrap prop below RSH
b <- boot::boot(
  data = df_obs$height,
  statistic = prop_between, stype = "i",
  R = 1000,
  lwr = 0, upp = round(min_rsh)
)

prop_below_height <- set_random("rnorm", mean = b$t0, sd = sd(b$t))

## bootstrap prop below max RSH
b <- boot::boot(
  data = df_obs$height,
  statistic = prop_between, stype = "i",
  R = 1000,
  lwr = 0, upp = round(max_rsh)
)

prop_below_max <- set_random("rnorm", mean = b$t0, sd = sd(b$t))

# defining prop_at_height as prop_below_max-prop_below_height ensures
# that prop_at_height + prop_below_height will never be greater than 1 
```

#### Encounter Rate

```{r}
setDT(df_obs)
setDT(df_survey)

df_obs_survey <- df_obs[, .("size" = sum(size)), survey_id][df_survey,
                                                            on = "survey_id"]

setnafill(df_obs_survey, cols = c("size"), fill = 0) # set surveys without observations to 0 
```

Now we can use this table in `encounter_rate()`:

```{r}

er_fun <- function(dat, i){
  return(encounter_rate(df_obs_summary = dat[i]))
}

er_boot <- boot::boot(data = df_obs_survey
                , statistic = er_fun,
                , sim = "ordinary"
                , stype = "i"
                , R = 200
)


## Define Encounter rate
 
encounter_rate <- set_random("rnorm",
                  mean = er_boot$t0,
                  sd = sd(er_boot$t, na.rm = TRUE)
                  )

## take a look
## 
encounter_rate_hist <- sapply(1:1000, FUN = function(i) collision::sample_input(encounter_rate) )
hist(encounter_rate_hist)
```

## Define a set of turbines

```{r}
df_turbines <- data.frame(
  turbine_id = c("T01", "T02"),
  lat = c(-32.512, -32.521),
  lon = c(143.441, 143.442)
)
```

### Run the simulation

```{r}
## inputs needed:
## seed for reproducibility
## iterations - how many runs
## df_turbines  - class `data.frame`
## wte   - class `birdInput`
## v90_single - class `turbineInput`
## prop_at_height - class `randomInput`
## prop_below_height - class `randomInput`
## edr - class `randomInput`
## observations - df_obs_survey


set.seed(1234)
iterations <- 1000

lst_results <- lapply(1:iterations, function(i) {
  df_turbines_results <- df_turbines

  # Step 1 - Sample inputs for bird, turbine, prop at height and prop below height
  bird_i <- sample_input(wte)
  turbines_i <- sample_input(v90_single)
  prop_below_height_i <- sample_input(prop_below_height)
  prop_below_max_i <- sample_input(prop_below_max)
  prop_at_height_i <- prop_below_max_i-prop_below_height_i
  edr_i <- sample_input(edr)
  er_i <- sample_input(encounter_rate)
  mean_h_i <- sample_input(mean_h)

  
  # Step 2 calculate flux through turbine per year
  df_turbines_results$flight_turbine_year <- turbine_flights_year(
    survey_type = c("point"),
    encounter_rate = er_i,
    eff_detection_width = 2*edr_i,
    mean_flight_height = mean_h_i,
    rotor_diameter = turbines_i$rotor_diam,
    hub_height = turbines_i$hh,
    prop_day = bird_i$prop_day,
    prop_year = bird_i$prop_year
  )

  # Step 4 - Run `p_collisions()` - calculate $P(C|I)$ for one interaction
  df_turbines_results$p_coll_static <- prob_collision_static(
    d_base = turbines_i$d_base,
    d_rotormin = turbines_i$d_rotormin,
    d_top = turbines_i$d_top,
    hh = turbines_i$hh,
    blade_length = turbines_i$blade_length,
    max_nac_h = turbines_i$max_nac_h,
    max_nac_l = turbines_i$max_nac_l,
    max_width_nacelle = turbines_i$max_width_nacelle,
    rotor_diam = turbines_i$rotor_diam,
    tilt_deg = turbines_i$tilt_deg,
    max_chord = turbines_i$max_chord,
    min_chord = turbines_i$min_chord,
    blade_thickness_wide = turbines_i$blade_thickness_wide,
    blade_thickness_narrow = turbines_i$blade_thickness_narrow,
    prop_at_height = prop_at_height_i,
    prop_below_height = prop_below_height_i
  )

  df_turbines_results$p_coll_dyn <- prob_collision_dynamic(
    rpm = turbines_i$rpm, # s_rot,
    blade_length = turbines_i$blade_length,
    max_width_nacelle = turbines_i$max_width_nacelle,
    rotor_diam = turbines_i$rotor_diam,
    blade_thickness_wide = turbines_i$blade_thickness_wide,
    blade_thickness_narrow = turbines_i$blade_thickness_narrow,
    hh = turbines_i$hh,
    bird_length = bird_i$bird_length,
    bird_speed = bird_i$bird_speed,
    prop_at_height = prop_at_height_i,
    prop_below_height = prop_below_height_i
  )


  # Step 5 - Run `n_collisions()` - calculate number of collisions a year
  df_turbines_results$n_collision <- n_collision(
    avoidance_rate_static = bird_i$avoidance_static,
    avoidance_rate_dynamic = bird_i$avoidance_dynamic,
    n_flights = df_turbines_results$flight_turbine_year,
    p_coll_static = df_turbines_results$p_coll_static,
    p_coll_dynamic = df_turbines_results$p_coll_dyn
  )

  ## for reporting - optional
  df_turbines_results$iteration <- i

  return(df_turbines_results)
})
```

Now we have the output of each simulation. We can use them to plot out the histogram of collision results.

```{r}
lapply(lst_results, function(x) {
  sum(x$n_collision)
}) |>
  unlist() |>
  hist(x = _, xlab = "n_collision", main = "Histogram of collision results")
```


