---
title: "Simple simulation example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simple simulation example}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE
)
```

```{r setup, results='hide'}
library(collision)
library(Distance)
library(mc2d)  # for the PERT distribution
```


# Stochastic Example

This is a good example to start with stochastic simulation. It models a (fictional) site with two proposed turbines, all of the same model, and a (fictional) point count survey of a raptor species. 

For this example we assume flight density is constant across the site.

We are going to calculate a simple stochastic model output here.  

In the examples below we are just outlining how to use the model but there's a lot of analysis required beforehand to fit a distance curve and extract the effective detection radius, analyse flight heights, and understand the spatial and temporal variability in the flight patterns. Please refer to other vignettes for more information.

## Setup inputs


### Define turbine model

We set up a turbine - here's an example based basically on a Vestas 90, set up using the `define_turbine` function. These parameters can be defined as probability distributions or as single numbers. 

```{r}
v90_single <- define_turbine(
  model_id = "Vesta V90",
  blade_length = 44,
  blade_thickness_narrow = 0.07,
  blade_thickness_wide = 2.6,
  d_base = 4.15,
  d_rotormin = 3.55,
  d_top = 2.55,
  hh = 65,
  max_chord = 3.51,
  min_chord = 0.39,
  max_nac_h = 4.05,
  max_nac_l = 13.25,
  max_width_nacelle = 3.6,
  rpm = 16.1,
  rotor_diam = NULL,
  tilt_deg = 6,
  prop_operational = 0.98
)
```

### Configure bird inputs

We make a 'bird' object - a list of key parameters. These parameters can be defined as probability distributions or as single numbers. 

```{r}
#TODO issue13 more info on distribuutions

wte <- define_bird(
  species = "Wedge-tailed Eagle",
  bird_length = set_random("rnorm", mean = 0.945, sd = 0.2),
  bird_speed = set_random("rnorm", mean = 17, sd = 3),
  prop_day = set_random("runif", min = 0.48, max = 0.52),
  prop_year = 1,
  avoidance_dynamic = set_random("rpert",
                                 min = 0.88, max = 0.95, 
                                 mode = 0.92, shape =4),
  avoidance_static = 0.9999
)


```


### Proportion of heights

We define the proportion of flights at rotor swept height and below rotor swept height. We will fit a distribution to our height data and extract the value at height

```{r}

collision::df_obs_survey



prop_at_height <- parameterise_lnorm(
  mean = 0.185,
  sd = 0.05
)

prop_at_height <- set_random(
  distr = "rlnorm",
  meanlog = prop_at_height$meanlog,
  sdlog = prop_at_height$sdlog
)

prop_below_height <- parameterise_lnorm(
  mean = 0.0432,
  sd = 0.01
)

prop_below_height <- set_random(
  distr = "rlnorm",
  meanlog = prop_below_height$meanlog,
  sdlog = prop_below_height$sdlog
)
```


## Define a set of turbines

The basic template for the model outputs is a `data.frame` of turbine inputs.  These can be read in from a csv, or you can set them up with a few basic pieces of information and the turbine definition above.  

For each turbine we need a turbine ID, and a location. Location can be NA if you are not including any spatial modelling.


```{r}
df_turbines <- data.frame(
  turbine_id = c("T01", "T02"),
  lat = c(-32.512, -32.521),
  lon = c(143.441, 143.442)
)
```


## Define a flight flux

For now, I'm assuming you have done the required distance correction and have arrived at a flight flux. This must be in units of **flights / min**.

I'm also assuming constant flight rate across the site.

Because we want to avoid negative flight flux, we're using a log-normal distribution. The right distribution requires analysis on a case-by-case basis.

```{r}
flight_flux_min <- parameterise_lnorm(
  mean = 0.000238,
  sd = 0.00001
)

flight_flux_min <- set_random(
  distr = "rlnorm",
  meanlog = flight_flux_min$meanlog,
  sdlog = flight_flux_min$sdlog
)
```

## Modelling

Here we set up a loop to run the simulation. Note that all the iteration of the loop are independent to others, meaning that you can wrap the loop inside methods to run the iterations in parallel to save time if you wish.

* Step 0 - Set up simulation parameters
* Step 1 - Sample inputs for bird, turbine, prop at height and prop below height
* Step 2 - Get sampled `flux_per_year()` - calculate flux through turbine per year
* Step 3 - Run `prob_interaction_flat()` - calculate $P(I)$
* Step 4 - Run `p_collisions()` - calculate $P(C|I)$ for one interaction
* Step 5 - Run `n_collisions()` - calculate number of collisions a year

For a breakdown of each step see [the deterministic examples](deterministic-example.html).  

### Run the simulation

```{r}
## inputs needed:
## seed for reproducibility
## iterations - how many runs
## df_turbines  - class `data.frame`
## wte   - class `birdInput`
## v90_single - class `turbineInput`
## prop_at_height - class `randomInput`
## prop_below_height - class `randomInput`
## flight_flux_min - class `randomInput`


set.seed(1234)
iterations <- 1000

lst_results <- lapply(1:iterations, function(i) {
  df_turbines_results <- df_turbines

  # Step 1 - Sample inputs for bird, turbine, prop at height and prop below height
  bird_i <- sample_input(wte)
  turbines_i <- sample_input(v90_single)
  prop_at_height_i <- sample_input(prop_at_height)
  prop_below_height_i <- sample_input(prop_below_height)

  # Step 2 - Get sampled `flux_per_year()` - calculate flux through turbine per year
  df_turbines_results$flight_flux_year <- flux_per_year(
    flux_per_min = sample_input(flight_flux_min),
    prop_day = bird_i$prop_day,
    prop_year = bird_i$prop_year
  )

  # Step 3 - Run `prob_interaction_flat()` - calculate $P(I)$
  df_turbines_results$prob_interaction <- prob_interaction_flat(num_turbines = nrow(df_turbines))


  # Step 4 - Run `p_collisions()` - calculate $P(C|I)$ for one interaction
  df_turbines_results$p_coll_static <- prob_collision_static(
    d_base = turbines_i$d_base,
    d_rotormin = turbines_i$d_rotormin,
    d_top = turbines_i$d_top,
    hh = turbines_i$hh,
    blade_length = turbines_i$blade_length,
    max_nac_h = turbines_i$max_nac_h,
    max_nac_l = turbines_i$max_nac_l,
    max_width_nacelle = turbines_i$max_width_nacelle,
    rotor_diam = turbines_i$rotor_diam,
    tilt_deg = turbines_i$tilt_deg,
    max_chord = turbines_i$max_chord,
    min_chord = turbines_i$min_chord,
    blade_thickness_wide = turbines_i$blade_thickness_wide,
    blade_thickness_narrow = turbines_i$blade_thickness_narrow,
    prop_at_height = prop_at_height_i,
    prop_below_height = prop_below_height_i
  )

  df_turbines_results$p_coll_dyn <- prob_collision_dynamic(
    rpm = turbines_i$rpm, # s_rot,
    blade_length = turbines_i$blade_length,
    max_width_nacelle = turbines_i$max_width_nacelle,
    rotor_diam = turbines_i$rotor_diam,
    blade_thickness_wide = turbines_i$blade_thickness_wide,
    blade_thickness_narrow = turbines_i$blade_thickness_narrow,
    hh = turbines_i$hh,
    bird_length = bird_i$bird_length,
    bird_speed = bird_i$bird_speed,
    prop_at_height = prop_at_height_i,
    prop_below_height = prop_below_height_i
  )


  # Step 5 - Run `n_collisions()` - calculate number of collisions a year
  df_turbines_results$n_collision <- n_collision(
    avoidance_rate_static = bird_i$avoidance_static,
    avoidance_rate_dynamic = bird_i$avoidance_dynamic,
    flux_yr = df_turbines_results$flight_flux_year,
    p_interaction = df_turbines_results$prob_interaction,
    p_coll_static = df_turbines_results$p_coll_static,
    p_coll_dynamic = df_turbines_results$p_coll_dyn
  )

  ## for reporting - optional
  df_turbines_results$iteration <- i

  return(df_turbines_results)
})
```

Now we have the output of each simulation. We can use them to plot out the hisogram of collision results.

```{r}
lapply(lst_results, function(x) {
  sum(x$n_collision)
}) |>
  unlist() |>
  hist(x = _, xlab = "n_collision", main = "Histogram of collision results")
```


